Spatio-angular fluorescence microscopy\\ I. Basic theory (Draft: \today)

### 2018-12-16 papers/basic-theory/paper/paper.tex
Rudolf's second round with discussion comments.

\title{Spatio-angular fluorescence microscopy I. Basic theory (Draft: \today)}

@@

\section{Introduction}
Fluorescence microscopes are widely used in the biological sciences for
measuring ensembles of fluorescent molecules that label specific proteins and
biologically important molecules. While most fluorescence microscopy experiments
are designed to measure only the spatial distribution of fluorophores, a growing
number of experiments seek to measure both the spatial and angular distributions
of fluorophores by the use of polarizers \cite{vrabioiu2006, mattheyses2010,
  mehta2016, mcquilken2017, zhanghao2017} or point spread function engineering
\cite{agrawal2012, zhang2018}.

@@

We use the term \textit{monopole} to refer to a model of a fluorophore that
treats it as an isotropic absorber/emitter. Although the term \textit{monopole
  approximation} is not in widespread use, we think it accurately describes the
way many models of fluorescence microscopes treat fluorophores, and we use the
term to distinguish the monopole model from more realistic dipole and
higher-order models. Despite their use in models, electromagnetic monopole
absorber/emitters do not exist in nature. All physical fluorophores absorb and
emit radiation with dipole or higher-order moments, and these moments are always
oriented in space. For a classical mental model of fluorophores we imagine each
dipole as a small oriented antenna (with incoherent absorption and emission
moments) where electrons are constrained to move along a single direction. With
this model we can argue classically that monopole radiators do not exist because
they would violate the conservation of charge.

Although many works implicitly apply the monopole approximation, we have
encountered two explicit justifications: (1) the sample contains many randomly
oriented fluorophores within a resolvable volume or (2) the sample contains
unconstrained rotating fluorophores. While both of these situations yield
monopole-like samples, neither yields a sample that is perfectly described by
the monopole model. In this work we investigate the dipole model of fluorophores
in detail and find the conditions under which the monopole approximation is
justified.

@@

While the monopole approximation applies to the fluorescent object, the
\textit{scalar approximation} applies to the fields that propagate through the
microscope. Modeling the electric fields in a region requires a
three-dimensional vector field, but if the electric fields are random or
parallel a scalar field is sufficient, and we can replace the vector-valued electric
field, $\mb{E}$, with a scalar-valued field, $U$.%  is the scalar approximation,
% and this approximation is only justified if the fields are randomly or uniformly
% polarized.

The scalar approximation is often made together with the monopole approximation.
For example, the Born-Wolf model \cite{born1980} and the Gibson-Lanni model
\cite{gibson89} make both the monopole and scalar approximations when applied to
fluorescence microscopes. However, some models make the monopole approximation
but not the scalar approximation. For example, the Richards-Wolf model
\cite{richards} considers the role of vector-valued fields in the optical
system, but it is an optical model so when it is applied to fluorescence
microscopes the monopole approximation is assumed. % As previously mentioned,
% several works in the SMLM community have modeled and applied dipole point spread
% functions that do not make the monopole or scalar approximations, but to our
% knowledge no works have considered dipole transfer functions.

This work lies at the intersection of three subfields of fluorescence
microscopy: (1) spatial ensemble imaging where fluorophores are imaged as
ensembles and the goal is to find their concentration as a function of position
in the sample, (2) the spatio-angular ensemble imaging where fluorophores are
imaged as ensembles and the goal is to find their concentration and orientation
as a function of position in the sample, and (3) SMLM imaging where fluorophores
are sparse in the sample and the goal is to find the position and orientation of
each fluorophore. We briefly review how these three subfields use the monopole
and scalar approximations.

The large majority of fluorescence microscopes are used to image ensembles of
fluorophores, and all existing restoration techniques make either or both of the
monopole and scalar approximations. As discussed above, the Gibson-Lanni model,
the Born-Wolf model, and the Richards-Wolf model are approximate when applied to
fluorescence microscopy data because they only model monopole emitters.
Deconvolution algorithms that use these models may make biased estimates of
fluorophore concentrations since they ignore the dipole excitation and emission
of fluorophores, but we will see that these effects are small when the sample
contains many randomly oriented fluorophores within a resolvable volume or
unconstrained rotating fluorophores.

A small but growing group of microscopists is interested in measuring the
orientation and position of ensembles of fluorophores \cite{vrabioiu2006,
  mattheyses2010, mehta2016, mcquilken2017, zhanghao2017}. These techniques
typically use polarizers to make multiple measurements of the same object, then
they use a model of the dipole excitation and emission processes
\cite{fourkas2001} to recover the orientation of fluorophores using pixel-wise
arithmetic. Although these studies do not adopt the scalar or monopole
approximations for the angular part of the problem, they adopt both
approximations when they consider the spatial part of the problem. Existing
works either ignore the spatial reconstruction problem \cite{vrabioiu2006,
  mattheyses2010, mehta2016, mcquilken2017} or assume that the spatial and
angular reconstruction problems can be solved sequentially
\cite{zhanghao2017}.

@@

In this work we place these three subfields on a common theoretical footing. We
begin in section \ref{sec:theory} by formulating the imaging problem
mathematically. We start with an abstract description of fluorescence imaging
systems and extend the usual monopole imaging model to dipoles. Along the way we
introduce transfer functions that will allow us to efficiently analyze
spatio-angular microscopes. In section \ref{sec:results} we use the mathematical
tools we developed in section \ref{sec:theory} to analyze and simulate a
single-view fluorescence microscope, and we demonstrate our techniques with four
simple phantoms. Finally, in section \ref{sec:discussion} we discuss the results
and their broader implications.

@@

We can plug Eq. \eqref{eq:absquarescalar} into Eq. \eqref{eq:otf} and use the
autocorrelation theorem to rewrite the monopole transfer function as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\taup\, C(\taup)C^*(\taup - \bv), 
\end{align}
where we have introduced the
\textit{monopole coherent transfer function} as the two-dimensional Fourier
transform of the monopole coherent spread function:
\begin{align}
  C(\taup) = \int_{\mbb{R}^2}d\mb{r}\, c(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\taup].
\end{align}
Physically, the monopole coherent transfer function corresponds to the
scalar-valued field in a Fourier plane of the detector with appropriate scaling.
% We have chosen the symbol $\taup$ as the coordinate for the coherent
% transfer function because the Fourier plane of the detector is often called the
% \textit{pupil plane}.

@@

 Backer and Moerner \cite{backer2014} have calculated the dipole pupil function
 for a high-NA objective as
 \begin{align}
   \mb{p}(\rp, \so) \propto
   % \begin{bmatrix}
   %   \tilde{C}_0\left(\frac{r_p}{f_o}\right) + \tilde{C}_2\left(\frac{r_p}{f_o}\right)\cos(2\phi_p)&\tilde{C}_2\left(\frac{r_p}{f_o}\right)\sin(2\phi_p)&\tilde{C}_1\left(\frac{r_p}{f_0}\right)\cos\phi_p\\
   %   \tilde{C}_2\left(\frac{r_p}{f_o}\right)\sin(2\phi_p)&\tilde{C}_0\left(\frac{r_p}{f_o}\right) - \tilde{C}_2\left(\frac{r_p}{f_o}\right)\cos(2\phi_p)&\tilde{C}_1\left(\frac{r_p}{f_0}\right)\sin\phi_p\\     
   %   0&0&0\\     
   % \end{bmatrix}
   \begin{bmatrix}
     \tilde{C}_0\left(\frac{r_p}{f_o}\right) + \tilde{C}_2\left(\frac{r_p}{f_o}\right)c(2\phi_p)&\tilde{C}_2\left(\frac{r_p}{f_o}\right)s(2\phi_p)&\tilde{C}_1\left(\frac{r_p}{f_0}\right)c(\phi_p)\\
     \tilde{C}_2\left(\frac{r_p}{f_o}\right)s(2\phi_p)&\tilde{C}_0\left(\frac{r_p}{f_o}\right) - \tilde{C}_2\left(\frac{r_p}{f_o}\right)c(2\phi_p)&\tilde{C}_1\left(\frac{r_p}{f_0}\right)s(\phi_p)\\     
     0&0&0\\     
   \end{bmatrix}
   \begin{bmatrix}
     s_x\\
     s_y\\
     s_z
   \end{bmatrix}
   \Pi\left(\frac{r_p}{2f_os(\alpha)}\right),
 \end{align}
 where $c(x)$ and $s(x)$ are shorthand for $\cos(x)$ and $\sin(x)$,
 $\{s_x, s_y, s_z\}$ are the Cartesian components of $\so$ when $\mh{z}$ is
 aligned with the optical axis, and
 \begin{alignat}{4}
   \tilde{C}_0(x) &= \frac{1}{2}\left(\sqrt{1 - x^2} + 1\right)(1 - x^2)^{-1/4} &&= 1 + \frac{x^4}{32} + \frac{x^6}{32} + \cdots,\\
   \tilde{C}_1(x) &= x(1 - x^2)^{-1/4} &&= x + \frac{x^3}{4} + \frac{5x^5}{32} + \cdots,\\
   \tilde{C}_2(x) &= \frac{1}{2}\left(\sqrt{1 - x^2} - 1\right)(1 - x^2)^{-1/4} &&= -\frac{x^2}{4} - \frac{x^4}{8} - \frac{11x^6}{128} - \cdots.
 \end{alignat}
 Similar to the monopole case, the dipole pupil function conserves power and has
 a cutoff at the objective aperture, but the dipole pupil function is
 vector-valued to model the complete electric field in the pupil plane. The
 fields in the pupil plane have a negligible $\mh{z}$ component which is a
 consequence of our assumption that the tube lens is paraxial---modeling a
 non-paraxial tube lens would require a three-dimensional vector-valued pupil
 function \cite{sheppard1994, gu2000, arnison2002, foreman2011-2}.

@@

\begin{figure}[h]
 % \captionsetup{width=1.0\linewidth}
 \centering
   \centering
   \includegraphics[scale=0.8]{../figures/microscope/microscope.pdf}
   \caption{Comparison of paraxial models for monopole radiators a) and dipole
     radiators b)--d). a) Monopole radiators fill the pupil plane with a uniform
     scalar field which gives rise to an Airy disk on the detector. b) A
     transverse dipole radiator also creates an Airy disk, but the pupil plane
     is filled with a uniform vector field. c) An axial dipole radiator creates
     a radial electric field pattern in the back focal plane that creates a
     $\text{jinc}_1^2(r)$ pattern on the detector. d) Dipoles that are not
     transverse or axial still create radially symmetric irradiance patterns
     under the paraxial approximation. Fields from transverse dipoles are real
     and even while fields from axial dipoles are real and odd, which causes a
     relative $\pi/2$ phase shift for the fields on the detector. This phase
     shift means that the fields from transverse and axial components of the
     dipole do not interfere, which causes radially symmetric irradiance
     patterns.}
   \label{fig:microscopefig}
 \end{figure}

@@

 \begin{figure}[h]
 % \captionsetup{width=1.0\linewidth}
 \centering
   \centering
   \includegraphics[scale=0.8]{../figures/paratfs/ph1.pdf}
   \caption{\textbf{Left:} A spatially and angularly sparse phantom---uniformly
     spaced single dipoles with varying orientations (increasing $\vartheta$
     from left to right and increasing $\varphi$ from bottom to top). White
     crosses mark the positions of the dipoles. \textbf{Center:} Scaled
     irradiance for an imaging system with $\text{NA} = 0.75$,
     $\lambda = 500\,\text{nm}$, and $n_o = 1.33$ sampled at $20\times$ the
     Nyquist rate. \textbf{Right:} $x$ profiles through the scaled irradiance.
     The response is independent of the azimuth angle and strongly dependent on
     the inclination angle.}
   \label{fig:ph1}
 \end{figure}

@@

  \begin{figure}[h]
 % \captionsetup{width=1.0\linewidth}
 \centering
   \centering
   \includegraphics[scale=0.8]{../figures/paratfs/ph3.pdf}
   \caption{\textbf{Left:} A spatially sparse and angularly dense
     phantom---uniformly spaced double cone distributions of fluorophores with
     varying central direction (increasing $\vartheta'$ from left to right) and
     varying cone half-angle (increasing $\Delta$ from bottom to top).
     \textbf{Center:} Scaled irradiance for an imaging system with
     $\text{NA} = 0.75$, $\lambda = 500\,\text{nm}$, and $n_o = 1.33$ sampled at
     $20\times$ the Nyquist rate. \textbf{Right:} $x$ profiles through the
     scaled irradiance. Small cone angles have irradiance patterns that vary
     with the central direction, while larger cones angles have increasingly
     uniform irradiance patterns that hide angular information.}
   \label{fig:ph3}
 \end{figure}

@@

  \begin{figure}[h]
 % \captionsetup{width=1.0\linewidth}
 \centering
   \centering
   \includegraphics[scale=0.8]{../figures/paratfs/ph4.pdf}
   \caption{\textbf{Left:} A spatially and angularly dense phantom---uniformly
     spaced disks with varying size (increasing $D$ from bottom to top) and
     double cone half angle (increasing $\Delta$ from left to right)
     \textbf{Center:} Scaled irradiance for an imaging system with
     $\text{NA} = 0.75$, $\lambda = 500\,\text{nm}$, and $n_o = 1.33$ sampled at
     $20\times$ the Nyquist rate. \textbf{Right:} $x$ profiles through the
     scaled irradiance.}
   \label{fig:ph4}
 \end{figure}

@@

However, the spherical harmonics provide several advantages over the second
moments. First, the spherical harmonics form a complete basis for functions on
the sphere, while the second moments span a much smaller function space. The
usual approach to extending the span of the second moments is to use the fourth
(or higher) moments, but this extension requires a completely new set of basis
functions while the spherical harmonics can be extended by simply adding higher order
terms. Second, the spherical harmonics are orthonormal, which will allow us to
deploy invaluable tools from linear algebra---linear subspaces, rank,
decompositions, etc.--- to analyze and compare microscope designs. Finally,
using the spherical harmonics provides access to a set of fast algorithms. The
naive expansion of an arbitrary discretized $N$ point spherical function onto
spherical harmonics (or second moments) requires a $\mc{O}(N^2)$ matrix
multiplication, while pioneering work by Driscoll and Healy \cite{driscoll1994}
showed that the forward discrete fast spherical harmonic transform can be
computed with a $\mc{O}(N(\log N)^2)$ algorithm and its inverse can be computed
with a $\mc{O}(N^{3/2})$ algorithm. To our knowledge no similarly fast
algorithms exist for expansion onto the higher-order moments.

@@

\subsection{What determines the angular bandwidth?}
Spatial imaging systems have a spatial bandwidth that characterizes the highest
spatial frequency that the system can image. Similarly, angular imaging systems
have an angular bandwidth that characterizes the highest angular frequency that
the system can image, but in the angular case there are two different types of
angular bandwidths that we call the $\ell$- and $m$-bandwidth. The $\ell$-bandwidth
can be interpreted in a similar way to the spatial bandwidth---it characterizes
the smallest angular features that the imaging system can measure. The
$m$-bandwidth does not have a direct analog in the spatial domain---it
characterizes the angular uniformity of the imaging system. If the $\ell$- and
$m$-bandwidths are equal then the imaging system can be said to have an 
$\textit{isotropic angular bandwidth}$.

@@

The microscope we considered in this work has an $\ell$-bandwidth of $\ell_c = 2$
and an $m$-bandwidth of $m_c = 0$, so it does not have an isotropic angular
bandwidth. In future work we will consider several approaches to improving the
angular bandwidths in detail, but we briefly mention that non-paraxial
microscopes, microscopes with polarizers on the illumination or detection paths,
and multiview microscopes all have higher angular bandwidths than the microscope
considered here.

@@

\subsection{Towards more realistic models}
The theoretical model we presented in this work is an extreme simplification of
a real microscope. We have ignored the effects of thick samples,
refractive-index mismatch, aberration, scattering, finite fluorescence
lifetimes, and interactions between fluorophores among others. Because of this
long list of unknown effects, real experiments will likely require extensions of
the models developed here.

@@

\section{Spherical Fourier transform of a double cone}\label{sec:cone}
In this appendix we evaluate the spherical Fourier transform of a
normalized double-cone angular distribution with central direction $\mh{s}'$ and
cone half-angle $\Delta$
\begin{align}
  f_{\text{(cone)}}(\mh{s}; \mh{s}', \Delta) = \frac{1}{4\pi(1 - \cos\Delta)}\Pi\left(\frac{\mh{s}\cdot\mh{s}'}{2\cos\Delta}\right). \label{eq:doublecone}
\end{align}
The spherical Fourier transform is
\begin{align}
  F_{\ell\text{(cone)}}^m(\mh{s}', \Delta) = \int_{\mbb{S}^2}d\mh{s}\, f_{\text{(cone)}}(\mh{s}; \mh{s}', \Delta)Y_\ell^{m*}(\mh{s}).
\end{align}
% This integral will be difficult to evaluate unless we change coordinates to
% exploit the axis of symmetry $\mh{s}'$. The key idea is that
% $f_{\text{(cone)}}(\mh{s}; \mh{s}', \Delta)$ is a function of
% $\mh{s}\cdot\mh{s}'$ so it is a one dimensional function on the domain [-1,1]
% that can be expanded in terms of Legendre polynomials (a Fourier-Legendre
% series)
% \begin{align}
%   f_{\text{(cone)}}(\mh{s}\cdot\mh{s}', \Delta) = \sum_{\ell = 0}^{\infty}F_{\ell\text{(cone)}}(\mh{s}\cdot\mh{s}', \Delta)P_{\ell}(\mh{s}\cdot\mh{s}').
% \end{align}
% We can find the coefficients $F_{\ell\text{(cone)}}(\mh{s}\cdot\mh{s}', \Delta)$
% using Fourier's trick---multiply both sides by $P_{\ell}(\mh{s}\cdot\mh{s}')$,
% integrate over [-1,1], and exploit orthogonality
% $\int_{-1}^{1}dx\,P_{\ell}(x)P_{\ell'}(x) = \frac{2}{2\ell +
%   1}\delta_{\ell\ell'}$ to find that
% \begin{align}
%   F_{\ell\text{(cone)}}(x, \Delta) = \frac{2\ell + 1}{2}\int_{-1}^{1}dx\, f_{\text{(cone)}}(x, \Delta)P_\ell(x). \label{eq:legendre}
% \end{align}
% Plugging Eq. \eqref{eq:doublecone} into Eq. \eqref{eq:legendre} yields
% \begin{align}
%   F_{\ell\text{(cone)}}(x, \Delta) = \frac{2\ell + 1}{4\pi(1-\cos\Delta)}\int^{1}_{\cos\Delta}dx\, P_\ell(x). 
% \end{align}
% The integral evaluates to \cite[ch.~7.111]{gradshteyn2007}
% \begin{align}
%   \int_{\cos\Delta}^1 d\vartheta\, P_\ell(u) =
% \begin{cases}
%   1 - \cos\Delta\,, &\ell = 0\,,\\
%   \sin\Delta\, P_l^{-1}(\cos\Delta)\,, &\text{else},
% \end{cases}
% \end{align}
% where $P_l^{-1}(\cos\Delta)$ is the associated Legendre polynomial with order
% $m=-1$, not an inverse Legendre polynomial.

>>>>

\title{Spatio-angular fluorescence microscopy\\ I. Basic theory (Draft: \today)}

@@

\section{Introduction}
Fluorescence microscopes are widely used in the biological sciences for imaging
fluorescent molecules that label specific proteins and biologically important
molecules. While most fluorescence microscopy experiments are designed to
measure only the spatial distribution of fluorophores, a growing number of
experiments seek to measure both the spatial and angular distributions of
fluorophores by the use of polarizers \cite{vrabioiu2006, mattheyses2010,
  mehta2016, mcquilken2017, zhanghao2017} or point spread function engineering
\cite{agrawal2012, zhang2018}.

@@

We use the term \textit{monopole} to refer to a model of a fluorophore that
treats it as an isotropic absorber/emitter. Although the term \textit{monopole
  approximation} is not in widespread use, we think it accurately describes the
way many models of fluorescence microscopy treat fluorophores, and we use the
term to distinguish the monopole model from more realistic dipole and
higher-order models. Despite their use in models, electromagnetic monopole
absorber/emitters do not exist in nature. All physical fluorophores absorb and
emit radiation with dipole or higher-order moments, and these moments are always
oriented in space. For a classical mental model of fluorophores we imagine each
dipole as a small oriented antenna (with incoherent absorption and emission
moments) where electrons are constrained to move along a single direction. % With
% this model we can argue classically that monopole radiators do not exist because
% they would violate the conservation of charge.

Although many works implicitly apply the monopole approximation, we have
encountered two explicit justifications: (1) the sample contains many randomly
oriented fluorophores within a resolvable volume or (2) the sample contains
unconstrained rotating fluorophores. While both of these situations yield
monopole-like emitters, neither yields emitters that are perfectly described by
the monopole model. In this work we investigate the dipole model of fluorophores
in detail and find the conditions under which the monopole approximation is
justified.

@@

While the monopole approximation applies to the fluorescent object, the
\textit{scalar approximation} applies to the fields that propagate through the
microscope. Modeling the electric fields in a region requires a
three-dimensional vector field, but if the electric fields are random or
completely parallel, a scalar field is sufficient, and we can replace the
vector-valued electric field, $\mb{E}$, with a scalar-valued field, $U$.

The scalar approximation is often made together with the monopole approximation.
For example, the Born-Wolf model \cite{born1980} and the Gibson-Lanni model
\cite{gibson89} make both the monopole and scalar approximations when applied to
fluorescence microscopes. However, some models make the monopole approximation
but not the scalar approximation. For example, the Richards-Wolf model
\cite{richards} considers the role of vector-valued fields in the optical
system, but it is an optical model so when it is applied to fluorescence
microscopes the monopole approximation is assumed.

This work lies at the intersection of three subfields of fluorescence
microscopy: (1) spatial ensemble imaging where each resolvable volume contains
many fluorophores and the goal is to find the concentration of fluorophores as a
function of position in the sample, (2) spatio-angular ensemble imaging where
each resolvable volume contains many fluorophores and the goal is to find the
concentration and orientation of fluorophores as a function of position in the
sample, and (3) SMLM imaging where fluorophores are sparse in the sample and the
goal is to find the position and orientation of each fluorophore. We briefly
review how these three subfields use the monopole and scalar approximations.

The large majority of fluorescence microscopes are used to image ensembles of
fluorophores, and most existing modeling techniques make use of the monopole
approximation, the scalar approximation, or both. As discussed above, the
Gibson-Lanni model, the Born-Wolf model, and the Richards-Wolf model are
approximate when applied to fluorescence microscopy data because they only model
monopole emitters. Deconvolution algorithms that use these models may make
biased estimates of fluorophore concentrations since they ignore the dipole
excitation and emission of fluorophores, but we will see that these effects are
small when the sample contains many randomly oriented fluorophores within a
resolvable volume or unconstrained rotating fluorophores.

A small but growing group of microscopists is interested in measuring the
prevailing orientation and position of ensembles of fluorophores
\cite{vrabioiu2006, mattheyses2010, mehta2016, mcquilken2017, zhanghao2017}.
These techniques typically use polarizers to make multiple measurements of the
same object with different polarizer orientations, then they use a model of the
dipole excitation and emission processes \cite{fourkas2001} to recover the
orientation of fluorophores using pixel-wise arithmetic. Although these studies
do not adopt the scalar or monopole approximations for the angular part of the
problem, they adopt both approximations when they consider the spatial part of
the problem. Existing works either ignore the spatial reconstruction problem
\cite{vrabioiu2006, mattheyses2010, mehta2016, mcquilken2017} or assume that the
spatial and angular reconstruction problems can be solved sequentially
\cite{zhanghao2017}.

@@

In the present work we begin to place these three subfields on a common
theoretical footing. First, in section \ref{sec:theory}, we assemble the
mathematical tools for formulating the imaging problem. We start with an
abstract description of fluorescence imaging systems and extend the usual
monopole imaging model to dipoles. Along the way we introduce transfer functions
that will allow us to efficiently analyze spatio-angular microscopes. In section
\ref{sec:results} we analyze and simulate a single-view fluorescence microscope,
and we demonstrate our techniques with four simple phantoms. Finally, in section
\ref{sec:discussion} we discuss the results and their broader implications.

In this paper we focus on modeling an extremely simple optical system---a
single-view fluorescence microscope without polarizers. In future papers of this
series we will extend our models to include polarizers and multi-view
microscopes. Additionally, we have restricted this paper to the forward
problem---the mapping between a known object and the data. In future papers we
will consider the inverse problem, and the singular value decomposition (SVD)
will play a central role.

@@

We can plug Eq. \eqref{eq:absquarescalar} into Eq. \eqref{eq:otf} and use the
autocorrelation theorem to rewrite the monopole transfer function as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\taup\, C(\taup)C^*(\taup - \bv), 
\end{align}
where we have introduced the
\textit{monopole coherent transfer function} as the two-dimensional Fourier
transform of the monopole coherent spread function:
\begin{align}
  C(\taup) = \int_{\mbb{R}^2}d\mb{r}\, c(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\taup].
\end{align}
Physically, the monopole coherent transfer function corresponds to the
scalar-valued field in a Fourier plane of the detector with appropriate scaling.

@@

 Backer and Moerner \cite{backer2014} have calculated the dipole pupil function
 for a high-NA objective as
 \begin{align}
   \mb{p}(\rp, \so) \propto
   \begin{bmatrix}
     \tilde{C}_0\left(\frac{r_p}{f_o}\right) + \tilde{C}_2\left(\frac{r_p}{f_o}\right)c(2\phi_p)&\tilde{C}_2\left(\frac{r_p}{f_o}\right)s(2\phi_p)&\tilde{C}_1\left(\frac{r_p}{f_0}\right)c(\phi_p)\\
     \tilde{C}_2\left(\frac{r_p}{f_o}\right)s(2\phi_p)&\tilde{C}_0\left(\frac{r_p}{f_o}\right) - \tilde{C}_2\left(\frac{r_p}{f_o}\right)c(2\phi_p)&\tilde{C}_1\left(\frac{r_p}{f_0}\right)s(\phi_p)\\     
     0&0&0\\     
   \end{bmatrix}
   \begin{bmatrix}
     s_x\\
     s_y\\
     s_z
   \end{bmatrix}
   \Pi\left(\frac{r_p}{2f_os(\alpha)}\right),
 \end{align}
 where $c(x)$ and $s(x)$ are shorthand for $\cos(x)$ and $\sin(x)$,
 $\{s_x, s_y, s_z\}$ are the Cartesian components of $\so$ when $\mh{z}$ is
 aligned with the optical axis, and
 \begin{alignat}{4}
   \tilde{C}_0(x) &= \frac{1}{2}\left(\sqrt{1 - x^2} + 1\right)(1 - x^2)^{-1/4} &&= 1 + \frac{x^4}{32} + \frac{x^6}{32} + \cdots,\\
   \tilde{C}_1(x) &= x(1 - x^2)^{-1/4} &&= x + \frac{x^3}{4} + \frac{5x^5}{32} + \cdots,\\
   \tilde{C}_2(x) &= \frac{1}{2}\left(\sqrt{1 - x^2} - 1\right)(1 - x^2)^{-1/4} &&= -\frac{x^2}{4} - \frac{x^4}{8} - \frac{11x^6}{128} - \cdots.
 \end{alignat}
 Similar to the monopole case, the dipole pupil function conserves power and has
 a cutoff at the objective aperture, but the dipole pupil function is
 vector-valued to model the complete electric field in the pupil plane. The
 fields in the pupil plane have a negligible $\mh{z}$ component which is a
 consequence of our assumption that the tube lens is paraxial---modeling a
 non-paraxial tube lens would require a three-dimensional vector-valued pupil
 function \cite{sheppard1994, gu2000, arnison2002, foreman2011-2}.

@@

\begin{figure}[h]
 \centering
   \centering
   \includegraphics[scale=0.8]{../figures/microscope/microscope.pdf}
   \caption{Comparison of paraxial models for monopole radiators a) and dipole
     radiators b)--d). a) Monopole radiators fill the pupil plane with a uniform
     scalar field which gives rise to an Airy disk on the detector. b) A
     transverse dipole radiator also creates an Airy disk, but the pupil plane
     is filled with a uniform vector field. c) An axial dipole radiator creates
     a radial electric field pattern in the back focal plane that creates a
     $\text{jinc}_1^2(r)$ pattern on the detector. d) Dipoles that are not
     transverse or axial still create radially symmetric irradiance patterns
     under the paraxial approximation. Fields from transverse dipoles are real
     and even while fields from axial dipoles are real and odd, which causes a
     relative $\pi/2$ phase shift for the fields on the detector. This phase
     shift means that the fields from transverse and axial components of the
     dipole do not interfere, which causes radially symmetric irradiance
     patterns.}
   \label{fig:microscopefig}
 \end{figure}

@@

 \begin{figure}[h]
 \centering
   \centering
   \includegraphics[scale=0.8]{../figures/paratfs/ph1.pdf}
   \caption{\textbf{Left:} A spatially and angularly sparse phantom---uniformly
     spaced single dipoles with varying orientations (increasing $\vartheta$
     from left to right and increasing $\varphi$ from bottom to top). White
     crosses mark the positions of the dipoles. \textbf{Center:} Scaled
     irradiance for an imaging system with $\text{NA} = 0.75$,
     $\lambda = 500\,\text{nm}$, and $n_o = 1.33$ sampled at $20\times$ the
     Nyquist rate. \textbf{Right:} $x$ profiles through the scaled irradiance.
     The response is independent of the azimuth angle and strongly dependent on
     the inclination angle.}
   \label{fig:ph1}
 \end{figure}

@@

  \begin{figure}[h]
 \centering
   \centering
   \includegraphics[scale=0.8]{../figures/paratfs/ph3.pdf}
   \caption{\textbf{Left:} A spatially sparse and angularly dense
     phantom---uniformly spaced double cone distributions of fluorophores with
     varying central direction (increasing $\vartheta'$ from left to right) and
     varying cone half-angle (increasing $\Delta$ from bottom to top).
     \textbf{Center:} Scaled irradiance for an imaging system with
     $\text{NA} = 0.75$, $\lambda = 500\,\text{nm}$, and $n_o = 1.33$ sampled at
     $20\times$ the Nyquist rate. \textbf{Right:} $x$ profiles through the
     scaled irradiance. Small cone angles have irradiance patterns that vary
     with the central direction, while larger cones angles have increasingly
     uniform irradiance patterns that hide angular information.}
   \label{fig:ph3}
 \end{figure}

@@

  \begin{figure}[h]
 \centering
   \centering
   \includegraphics[scale=0.8]{../figures/paratfs/ph4.pdf}
   \caption{\textbf{Left:} A spatially and angularly dense phantom---uniformly
     spaced disks with varying size (increasing $D$ from bottom to top) and
     double cone half angle (increasing $\Delta$ from left to right)
     \textbf{Center:} Scaled irradiance for an imaging system with
     $\text{NA} = 0.75$, $\lambda = 500\,\text{nm}$, and $n_o = 1.33$ sampled at
     $20\times$ the Nyquist rate. \textbf{Right:} $x$ profiles through the
     scaled irradiance.}
   \label{fig:ph4}
 \end{figure}

@@

However, the spherical harmonics provide several advantages over the second
moments. First, the spherical harmonics form a complete basis for functions on
the sphere, while the second moments span a much smaller function space. The
usual approach to extending the span of the second moments is to use the fourth
(or higher) moments, but this extension requires a completely new set of basis
functions while the spherical harmonics can be extended by simply adding higher
order terms. Second, the spherical harmonics are orthonormal, which will allow
us to deploy invaluable tools from linear algebra---linear subspaces, rank, SVD,
etc.--- to analyze and compare microscope designs. Finally, using the spherical
harmonics provides access to a set of fast algorithms. The naive expansion of an
arbitrary discretized $N$ point spherical function onto spherical harmonics (or
second moments) requires a $\mc{O}(N^2)$ matrix multiplication, while pioneering
work by Driscoll and Healy \cite{driscoll1994} showed that the forward discrete
fast spherical harmonic transform can be computed with a $\mc{O}(N(\log N)^2)$
algorithm and its inverse can be computed with a $\mc{O}(N^{3/2})$ algorithm. To
our knowledge no similarly fast algorithms exist for expansion onto the
higher-order moments.

@@

\subsection{What determines the angular bandwidth?}
Spatial imaging systems have a spatial bandwidth that characterizes the highest
spatial frequency that the system can transfer between object and data space.
Similarly, angular imaging systems have an angular bandwidth that characterizes
the highest angular frequency that the system can transfer, but in the angular case
there are two different types of angular bandwidths that we call the $\ell$- and
$m$-bandwidth. The $\ell$-bandwidth can be interpreted in a similar way to the
spatial bandwidth---it characterizes the smallest angular features that the
imaging system can measure. The $m$-bandwidth does not have a direct analog in
the spatial domain---it characterizes the angular uniformity of the imaging
system. If the $\ell$- and $m$-bandwidths are equal then the imaging system can
be said to have an $\textit{isotropic angular bandwidth}$.

@@

The microscope we considered in this work has an $\ell$-bandwidth of
$\ell_c = 2$ and an $m$-bandwidth of $m_c = 0$, so it does not have an isotropic
angular bandwidth. In future work we will consider several approaches to
improving the angular bandwidths in detail, but we briefly mention that
non-paraxial microscopes, microscopes with polarizers in the illumination or
detection paths, and multiview microscopes all have higher angular bandwidths
than the microscope considered here.

@@

\subsection{Towards more realistic models}
The theoretical model we presented in this work is an extreme simplification of
a real microscope. We have ignored the effects of thick samples,
refractive-index mismatch, aberration, scattering, finite fluorescence
lifetimes, and interactions between fluorophores among others. Because of this
long list of unconsidered effects, real experiments will likely require
extensions of the models developed here.

@@

\section{Spherical Fourier transform of a double cone}\label{sec:cone}
In this appendix we evaluate the spherical Fourier transform of a
normalized double-cone angular distribution with central direction $\mh{s}'$ and
cone half-angle $\Delta$
\begin{align}
  f_{\text{(cone)}}(\mh{s}; \mh{s}', \Delta) = \frac{1}{4\pi(1 - \cos\Delta)}\Pi\left(\frac{\mh{s}\cdot\mh{s}'}{2\cos\Delta}\right). \label{eq:doublecone}
\end{align}
The spherical Fourier transform is
\begin{align}
  F_{\ell\text{(cone)}}^m(\mh{s}', \Delta) = \int_{\mbb{S}^2}d\mh{s}\, f_{\text{(cone)}}(\mh{s}; \mh{s}', \Delta)Y_\ell^{m*}(\mh{s}).
\end{align}
